More on the Slavnov products of spin chains, KP hierarchy tau and Baker-Akhiezer functions

### 2025-05-15 draft.tex
work on section 3 of the paper

Generically, we write the \(\mathcal{Y}\)-functions as 
\begin{equation}
  \mathcal{Y}(z; \bm{v}) = \sum_{p=0}^n \alpha_p(z) \sigma_p^{(n)}(\bm{v})\; , 
\end{equation}
where \(\sigma_p^{(n)}(\bm{v})\) are elementary symmetric polynomials
in \(\bm{v}\), and \(\alpha_p(z)\) are free functional parameters.
The most important point for us is that the functions \(\alpha_p(z)\), 
and consequently the \(\mathcal{Y}\)-function, are regular in the limit
\(z \to v_j \in \bm{v}\). The XXX spin chain has been discussed
in~\cite{Belliard:2019bfz}, and the authors show that \(\alpha_p(z)\)
are polynomials of degree \(2n\).

@@

We now define the matrix \(\bm{\Omega}\) by specifying its components
\begin{equation}
  \Omega_{jk}(u_k; \bm{v}) = g(u_k, v_j) \mathcal{Y}(u_k; \{u_k, \bm{v}_j\})\; ,
  \quad j= 1, \dots, n\; , \quad k =1, \dots, n+1\; .
\end{equation}
The matrices \(\hat{\Omega}_\ell\) are minors of
\(\bm{\Omega}\) with the \(\ell\)-th column excluded, in other words,
\begin{equation}
 \hat{\Omega}_\ell = \det_{k\neq \ell} \Omega_{jk}\; . 
\end{equation}

All in all, we use these objects to define the \(n+1\) functions below
\begin{equation}
  \label{eq:slavnov-tau}
  \bm{\tau}_\ell(\bm{u}_\ell; \bm{v}) \equiv \frac{1}{c_0}\frac{\bm{\zeta}_\ell}{\phi(\bm{v})}
  = \frac{\hat{\Omega}_{\ell}}{\Delta(\bm{u}_\ell)} \; .
\end{equation}
This is the most important result for our discussion; and our goal now
is to show that it is a tau-function of the KP-hierarchy.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Tau functions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In order to proceed, let us simplify the notation and organize the
results we described before. First of all, we need to observe
that we can define \(n\) functions \(\Omega_j(z)\) of the form
\begin{equation}
\begin{split}
  \Omega_{j}(z; \bm{v}) & = g(z, v_j) \mathcal{Y}(z; \{z, \bm{v}_j\})\qquad j =1, \dots, n\\
   & = \frac{c}{z - v_j} \mathcal{Y}(z; \{z, \bm{v}_j\})\; .
\end{split}
\end{equation}
Moreover, we have the expansion 
\begin{equation}
  \mathcal{Y}(z; \{z,\bm{v}_j\}) = \sum_{p=0}^n \alpha_p(z) \sigma_p^{(n)}(\{z,\bm{v}_j\})\; , 
\end{equation}
and using that the elementary symmetric polynomials decompose as
\begin{equation}
  \sigma_p^{(n)}(\{z,\bm{v}_j\}) = \sigma_p^{(n-1)}(\bm{v}_j) + z \sigma_{p-1}^{(n-1)}(\bm{v}_j) \; . 
\end{equation}

Let us also denote the functions
\begin{equation}
  \label{eq:y-functions-alpha}
  \mathcal{Y}_j(z)  \equiv  \mathcal{Y}(z; \{z,\bm{v}_j\})  = 
\sum_{p=0}^{n} \alpha_p(z)
  \left(\sigma_p^{(n-1)}(\bm{v}_j) + z \sigma_{p-1}^{(n-1)}(\bm{v}_j) \right) \; .
\end{equation}
Observe that the defining property of these functions
\(\mathcal{Y}_j(z)\) is the absence of the \(j\)-th Bethe root
\(v_j\).
Using that the elementary symmetric polynomials also satisfy \(\sigma_n^{(n-1)}(\bm{v}_j)=0\)
and \(\sigma_{-1}^{(n)}(\bm{v}_j)= 0\), we can write
\begin{equation}
  \label{eq:y-functions}
  \mathcal{Y}_j(z)  = 
  \sum_{p=0}^{n-1} 
   \left(\alpha_p(z) + z \alpha_{p+1}(z) \right) \sigma_p^{(n-1)}(\bm{v}_j) \; .
\end{equation}

Putting all these facts together, we have the functions
\begin{equation}
  \Omega_{j}(z; \bm{v}) = \frac{c}{z - v_j} \mathcal{Y}_j(z) \; .
\end{equation}
It is also convenient to consider \(c=1\), consequently \(c_0=1\).

\subsection{Alternant expression for the Slavnov products}

The matrix \(\bm{\Omega}\) can be written as
\begin{equation}
  \bm{\Omega} =
  \begin{pmatrix}
    \Omega_1(u_1) & \dots & \Omega_1(u_n) & \Omega_1(u_{n+1}) \\
    \Omega_2(u_1) & \dots & \Omega_2(u_n) & \Omega_2(u_{n+1}) \\
    \vdots & \vdots & \vdots & \vdots \\
    \Omega_n(u_1) & \dots & \Omega_n(u_n) & \Omega_n(u_{n+1})\; .
  \end{pmatrix}
\end{equation}
From this expression, we can also define a set of \(n\) square matrices
\(\bm\Omega^{(\ell)}(\bm{u}_\ell; \bm{v})\),
\(\ell = 1, \dots, n+1\), removing the \(\ell\)-th column of
\(\bm{\Omega}\), that is
\begin{equation}
  \bm{\Omega}^{(\ell)} =
  \begin{pmatrix}
    \Omega_1(u_1) & \dots & \Omega_1(u_{\ell-1}) & \Omega_1(u_{\ell+1})&\dots & \Omega_1(u_{n+1}) \\
    \vdots & \dots & \vdots & \vdots & \dots & \vdots \\
    \Omega_n(u_1) & \dots & \Omega_n(u_{\ell-1}) & \Omega_n(u_{\ell+1})&\dots & \Omega_n(u_{n+1})
  \end{pmatrix}\; .
\end{equation}

Finaly, we can write~(\ref{eq:slavnov-tau}) as 
\begin{equation}
\label{eq:slavnov-tau-functions}
  \bm{\tau}_\ell(\bm{u}_\ell; \bm{v}) = \frac{\det [\Omega_j((\bm{u}_\ell)_k)]|_{j,k=1}^n }{\Delta(\bm{u}_\ell)} 
\end{equation}
where \((\bm{u}_\ell)_k \) is the \(k\)-th component of the vector
\(\bm{u}_\ell\).  Therefore, for a fixed \(\ell\), we define the
vector \( \bm{z} \equiv \bm{u}_\ell\).  Consequently,
\begin{equation}
\begin{split}
\label{eq:slav-tau-function}
  \bm{\tau}(\bm{z}, \bm{v})
  & = \frac{1}{\Delta(\bm{z})} \det[\Omega_j(z_k) ]_{j,k=1}^n \\
  & = \frac{1}{\Delta(\bm{z})} \det\left[\frac{\mathcal{Y}_j(z_k)}{z_k - v_j} \right]_{j,k=1}^n  \; .
\end{split}
\end{equation}
All these tau functions have the form of an \emph{alternant
determinant} devided by the Vandermonde determinant.

It is well documented that the functions defined as
in~(\ref{eq:slav-tau-function}) are tau functions of the
KP-hierachy, see~\cite{Araujo:2021ghu} and references therein, in
particular~\cite{Kharchev:1991cy}. Consequently, these functions are
tau functions.

Henceforth, we will only refer to the matrix
\(\bm{\Omega}\) defined by the components
\begin{equation}
  \label{eq:omega-matrix}
  \Omega_{jk} = \frac{\mathcal{Y}_j(z_k)}{z_k - v_j}\; .
\end{equation}

\subsection{Tau functions expansion of the Slavnov product}

We can immediately observe that the poles of the tau
function~(\ref{eq:slav-tau-function}) are exactly the Bethe roots of
the integrable model.  We now want to analyse this pole structure of
\(\Delta(\bm{z})\tau(\bm{z}, \bm{v})\).  Let us consider that the
Bethe roots are not degenerate, that is, if \(j \neq k\), then \(v_j
\neq v_k\). In that case, the poles \(v_j\) are simple poles.

>>>>

Generically, we write the \(\mathcal{Y}\)-functions as 
\begin{equation}
\label{eq:yfunc-exp}
  \mathcal{Y}(z; \bm{v}) = \sum_{p=0}^n \alpha_p(z) \sigma_p^{(n)}(\bm{v})\; , 
\end{equation}
where \(\sigma_p^{(n)}(\bm{v})\) are elementary symmetric polynomials
in \(\bm{v}\), and \(\alpha_p(z)\) are free functional parameters.
The most important point for us is that the functions \(\alpha_p(z)\), 
and consequently the \(\mathcal{Y}\)-function, are regular in the limit
\(z \to v_j \in \bm{v}\). The XXX spin chain has been discussed
in~\cite{Belliard:2019bfz}, and the authors show that \(\alpha_p(z)\)
are polynomials of degree \(2n\).

@@

We now define the \(n\times (n+1)\) matrix \(\bm{\Omega}\) by specifying its components
\begin{equation}
  \label{eq:omega-matrix-def}
  \Omega_{jk}(u_k; \bm{v}) = g(u_k, v_j) \mathcal{Y}(u_k; \{u_k, \bm{v}_j\})\; ,
  \quad j= 1, \dots, n\; , \quad k =1, \dots, n+1\; .
\end{equation}
The matrices \(\hat{\Omega}_\ell\) are minors of
\(\bm{\Omega}\) with the \(\ell\)-th column excluded, in other words,
\begin{equation}
 \hat{\Omega}_\ell = \det_{k\neq \ell} \Omega_{jk}\; . 
\end{equation}
All in all, we use these determinants to define \(n+1\)
\emph{normalized Slavnov products} as follows
\begin{equation}
  \label{eq:slavnov-tau}
  \bm{\tau}^{(\ell)}(\bm{u}_\ell; \bm{v}) \equiv \frac{1}{c_0}\frac{\bm{\zeta}_\ell}{\phi(\bm{v})}
  = \frac{\hat{\Omega}_{\ell}}{\Delta(\bm{u}_\ell)}\; , \quad \ell = 1, \dots, n+1 \; .
\end{equation}
This is the most important result for our discussion; and our goal now
is to show that each \(\bm{\tau}^{(\ell)}(\bm{u}_\ell; \bm{v})\) is a
tau-function of the KP-hierarchy.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Tau functions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In order to proceed, let us simplify the notation and organize the
results described above. First of all, observe that we can define
\(n\) functions \(\Omega_j(z)\) of the form
\begin{equation}
\begin{split}
  \Omega_{j}(z) \equiv
  \Omega_{j}(z; \bm{v}) & = g(z, v_j) \mathcal{Y}(z; \{z, \bm{v}_j\})\\
   & = \frac{c}{z - v_j} \mathcal{Y}(z; \{z, \bm{v}_j\})\; , \qquad j =1, \dots, n \; .
\end{split}
\end{equation}
Let us also denote \(\mathcal{Y}(z; \{z, \bm{v}_j\}) \equiv
\mathcal{Y}_j(z)\), and note that we have \(n\) functions
\(\mathcal{Y}_j(z)\) whose defining property is the absence of the
\(j\)-th Bethe root \(v_j\).  It is also convenient to consider
\(c=1\), consequently \(c_0=1\).

Moreover, the expansion of the
\(\mathcal{Y}\)-functions~(\ref{eq:yfunc-exp}) yields
\begin{equation}
  \mathcal{Y}(z; \{z,\bm{v}_j\}) = \sum_{p=0}^n \alpha_p(z) \sigma_p^{(n)}(\{z,\bm{v}_j\})\; . 
\end{equation}
Using that the elementary symmetric polynomials satisfies the
relations
\begin{equation}
  \sigma_p^{(n)}(\{z,\bm{v}_j\}) = \sigma_p^{(n-1)}(\bm{v}_j) + z \sigma_{p-1}^{(n-1)}(\bm{v}_j) \; . 
\end{equation}
we write
\begin{equation}
  \label{eq:y-functions-alpha}
  \mathcal{Y}(z; \{z,\bm{v}_j\})  = 
\sum_{p=0}^{n} \alpha_p(z)
  \left(\sigma_p^{(n-1)}(\bm{v}_j) + z \sigma_{p-1}^{(n-1)}(\bm{v}_j) \right) \; .
\end{equation}
The elementary symmetric polynomials also satisfy
\(\sigma_n^{(n-1)}(\bm{v}_j)=0\) and \(\sigma_{-1}^{(n)}(\bm{v}_j)=
0\), and it yields
\begin{equation}
  \label{eq:y-functions}
  \mathcal{Y}_j(z)  = 
  \sum_{p=0}^{n-1} \beta_p(z) \sigma_p^{(n-1)}(\bm{v}_j) \; , \qquad 
\beta_p(z) = \alpha_p(z) + z \alpha_{p+1}(z) \; .
\end{equation}

Putting all these facts together, we have
\begin{equation}
\begin{split}
  \Omega_{j}(z; \bm{v}) & = \frac{\mathcal{Y}_j(z) }{z - v_j} \\ 
& = \frac{1}{z - v_j} \sum_{p=0}^{n-1} \beta_p(z) \sigma_p^{(n-1)}(\bm{v}_j) \; .
\end{split}
\end{equation}
The factor \((z - v_j)^{-1}\) introduces the dependence on the Bethe root \(v_j\).  
Moreover, one can observe that
\begin{equation}
\Res_{z=v_k} \left( \Omega_j(z)\right) =  \delta_{jk}\mathcal{Y}_j(v_j)\; .
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Alternant determinant expression for the Slavnov products}

Let us now write the matrix \(\bm{\Omega}\) defined
in~(\ref{eq:omega-matrix-def}) as
\begin{equation}
  \bm{\Omega} =
  \begin{pmatrix}
    \Omega_1(u_1) & \dots & \Omega_1(u_\ell) & \dots  & \Omega_1(u_n) & \Omega_1(u_{n+1}) \\
    \Omega_2(u_1) & \dots & \Omega_2(u_\ell) & \dots  & \Omega_2(u_n) & \Omega_2(u_{n+1}) \\
          \vdots &       &   \vdots    &    &     \vdots       & \vdots   \\
    \Omega_n(u_1) & \dots & \Omega_n(u_\ell)& \dots  & \Omega_n(u_n) & \Omega_n(u_{n+1})
  \end{pmatrix}\; .
\end{equation}
From this expression, we can also define \(n+1\) square matrices
\(\bm\Omega^{(\ell)}(\bm{u}_\ell) \equiv \bm\Omega^{(\ell)}(\bm{u}_\ell; \bm{v})\), for
\(\ell = 1, \dots, n+1\), by deleting the \(\ell\)-th column of
\(\bm{\Omega}\). That is
\begin{equation}
  \bm{\Omega}^{(\ell)} =
  \begin{pmatrix}
    \Omega_1(u_1) & \dots & \Omega_1(u_{\ell-1}) & \Omega_1(u_{\ell+1})&\dots & \Omega_1(u_{n+1}) \\
    \vdots &  & \vdots & \vdots & & \vdots \\
    \Omega_n(u_1) & \dots & \Omega_n(u_{\ell-1}) & \Omega_n(u_{\ell+1})&\dots & \Omega_n(u_{n+1})
  \end{pmatrix}\; .
\end{equation}
Henceforth, we will only refer to the matrix
\(\bm{\Omega}\) defined by the components
\begin{equation}
  \label{eq:omega-matrix}
  \Omega_{jk} = \frac{\mathcal{Y}_j(z_k)}{z_k - v_j}\; .
\end{equation}

Finally, we can write~(\ref{eq:slavnov-tau}) as 
\begin{equation}
\label{eq:slavnov-tau-functions}
  \bm{\tau}^{(\ell)}(\bm{u}_\ell; \bm{v}) = \frac{\det [\Omega_j((\bm{u}_\ell)_k)]|_{j,k=1}^n }{\Delta(\bm{u}_\ell)} \; ,
\end{equation}
where \((\bm{u}_\ell)_k \) is the \(k\)-th component of the vector
\(\bm{u}_\ell\).  More explicitly, remember that \(\bm{u} = (u_1,
\dots, u_{n+1})\) and that \(\bm{u}_\ell = \bm{u}_\ell\setminus
\{u_\ell\} = (u_1, \dots, u_{\ell-1}, u_{\ell+ 1}, \dots, u_{n+1})\),
we have
\begin{equation}
  (\bm{u}_\ell)_k =
  \left\{ 
  \begin{matrix}
    u_k & \text{if} & k < \ell\\
    u_{k+1} & \text{if} & k > \ell
  \end{matrix}
  \right. \; .
\end{equation}
It is convenient to write \(\bm{z}^{(\ell)} = (z^{(\ell)}_1,
z^{(\ell)}_2, \dots, z^{(\ell)}_n) = \bm{u}_{\ell}\). Moreover, it is
worth noting that the normalized Slavnov product
\(\bm{\tau}^{(\ell)}\) is completely independent of the parameter
\(u_\ell\). On the other hand, these functions are not independent from each
other, since \(\bm{z}^{(\ell)} \cap \bm{z}^{(\ell')} = \bm{u} \setminus
\{ u_\ell, u_{\ell'}\} \).

Consequently, we write the normalized Slavnov products as 
\begin{equation}
\begin{split}
\label{eq:slav-tau-function}
  \bm{\tau}^{(\ell)}(\bm{z}^{(\ell)}, \bm{v})
  & = \frac{1}{\Delta(\bm{z^{(\ell)}})} \det[\Omega_j(z^{(\ell)}_k) ]_{j,k=1}^n \\
  & = \frac{1}{\Delta(\bm{z}^{(\ell)})} \det\left[\frac{\mathcal{Y}_j(z^{(\ell)}_k)}{z^{(\ell)}_k - v_j} \right]_{j,k=1}^n
  \; \quad \ell = 1, \dots, n+1 \; .
\end{split}
\end{equation}
This is the first result of out work.

All the normalized Slavnov products~(\ref{eq:slav-tau-function}) take
the form of \emph{alternant determinants} of the functions
\(\{\Omega_j\}_{j=1}^n\) divided by Vandermonde determinants.  This
result establishes that the normalized slavnov
products~(\ref{eq:slav-tau-function}) are tau functions of the KP
hierarchy. Indeed, it is well documented that, given a set of generic
functions \(\{\phi_j\}_{j=1}^n\), expressions of the type
\begin{equation}
\tau(\bm{w}) =
\frac{\det_{i,j} \phi_i(w_j)}{\Delta(\bm{w})}\; ,
\end{equation}
where \(\{w_j\}_{j=1}^n\) are complex parameters, satisfy the Hirota
bilinear equation. Consequently, they are tau functions of the KP
hierarchy.  This proposition has been extensively discussed in the
literature, for example~\cite{Segal:1985aga, Kharchev:1991cy,
  Araujo:2021ghu} and references therein.

\todo{15/05}
\begin{remark}
  In~\cite{Segal:1985aga, Kharchev:1991cy, Alexandrov:2014cwa}, the
  authors have explored a family of functions \(\{ \phi_i\}_{i=1}^n\)
  that forms a basis vector because they parametrize points in a
  Grassmannian space. These functions are also useful in the
  definition of the Baker-Akhiezer function. The important aspect for
  us is that the functions \(\Omega_j\) relevant for our discussion do
  not belong to the class discussed in the references above.
  Therefore, we need to explore their properties and the
  Baker-Akhiezer functions relevant to our present discussion.
\end{remark}

\begin{remark}
  Reduction of a multicomponent KP Hierarchy
  It is interesting to observe that we have an array of tau functions ...
\end{remark}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tau functions expansion of the Slavnov product}
We can immediately observe that the poles of the tau
function~(\ref{eq:slav-tau-function}) are exactly the Bethe roots of
the integrable model.  We now want to analyse this pole structure of
\(\Delta(\bm{z})\tau(\bm{z}, \bm{v})\).  Let us consider that the
Bethe roots are not degenerate, that is, if \(j \neq k\), then \(v_j
\neq v_k\). In that case, the poles \(v_j\) are simple poles.
